---
title: "Advent of Code 2022 - Day 1"
output: github_document

---

```{r setup, include=FALSE}
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)

```

## Import and Tag Data

Load the data line by line as a character vector. Then convert to a data frame and add a row number column as a unique id for filtering later

```{r load_data}
input_data = readLines("input/input_1.txt")

input_df = data.frame(calories = input_data) %>% 
  mutate(row_number = row_number(), .before = 1)

```

## Identify the Groups

The rows with blank calories indicate the start of a new group. We can coerce calories to numeric and use is.na to locate the blanks for some boolean math
```{r flag_groups}

grouped_df = input_df %>% 
  mutate(
    calories = as.numeric(calories),
    is_blank = is.na(calories),
    elf_number = cumsum(is_blank) + 1
  ) 

```

## Summarize and Answer

Now that the elves are numbered, we can summarize by elf number to find the total calories for each elf.

Identifying the largest total calories and summing the top three is a simple max or sort-->slice-->sum.
```{r summarize}
total_calories = grouped_df %>% 
  summarise(
    .by = elf_number,
    total_calories = sum(calories, na.rm = TRUE)
  )

most_calories = total_calories %>% 
  pull(total_calories) %>% 
  max(na.rm = TRUE)

top_three_calories = total_calories %>% 
  arrange(desc(total_calories)) %>% 
  slice(1:3) %>% 
  pull(total_calories) %>% 
  sum()
```
